% !TEX root = ../discriminative_filtering.tex

I thank my parents for their continued love and support. I thank Prof. Matthew Harrison for his patience, Prof. Basilis Gidas for his dedication to graduate-level instruction, and Prof. Jerome Darbon for his willingness to discuss mathematics.  My office mates and rock climbing partners Michael Snarski, Ian Alevy, and Sameer Iyer have been great sources of encouragement.  Outside the office, Cat Munro, Clark Bowman, and Johnny Guzm\'an greatly enriched my life in Providence.  I'm grateful to our division's intramural soccer team (and our captains Dan Johnson and Guo-Jhen Wu) and hope that possibly without me they will manage a championship in the near future.  I am indebted to Dan Keating, and the entire Brown Polo Club family for teaching me the game of polo.  There is nothing quite so refreshing after a long day cooped up at a computer than riding around on a horse thwacking at grapefruit-sized ball.
   
My collaborators in neuroscience, particularly Dr. David Brandman and Prof. Leigh Hochberg, have given my work meaning.  Before coming to Brown, I had never imagined I would have the opportunity to work on such a practical and impactful project as BrainGate.  As a group, we are grateful to B. Travers, Y. Mironovas, D. Rosler, Dr. John Simeral, J. Saab, T. Hosman, D. Milstein, B. Shanahan, B. Sorice, J. Kelemen, B. Franco, B. Jarosiewicz, C. Vargas-Irwin, and the many others whose work has made this project possible over the years.  I was honored to help Chris Grimm complete his undergraduate thesis work in Bayesian filtering. Most importantly, we honor those who have dedicated some of the final years of their lives to participating in the project, especially T9 and T10, along with their families.
   
I thank the Brown--Kobe Exchange in High Performance Computing, and particularly Prof. Nobuyuki Kaya (\foreignlanguage{japanese}{\TenMincho\small 賀谷\;信幸}) for his generosity, the Brown--IMPA Partnership travel grant, and the Brown International and Conference travel awards for the opportunities to learn and share knowledge all around the world.  I also thank the Brown Institute for Brain Science graduate research award for a semester's respite from teaching.
   
I thank again those who championed me before coming to Brown, including Mrs. Kathleen Helbing who taught me the value and richness of history, Prof. Burgess Davis who first taught me analysis and probability, Prof. Robert Zink who introduced me to abstract algebra and served with me on the Purdue \emph{Exponent's} Board of Directors, Prof. Hans Uli Walther who guided me through my first research experience, Prof. Daniel Ocone who introduced me to stochastic differential equations, and Prof. Eduardo Sontag who guided my first graduate research experience.  I'm indebted to Prof. Victor Zavala for introducing me to Gaussian processes and piloting my first research paper.

In that vein, I'm grateful to John Wiegand and Chris O'Neil, and their families, with whom I went to church and elementary school, played soccer, went on adventures in Boy Scouts, and whom I continue to be fortunate to count as friends. John and the Wiegands have made coming home to Indiana something to look forward to.  The O'Neils were my first friends from Rhode Island, before I even knew what Rhode Island was.  I'm grateful to Ed Chien, my first climbing partner and a great friend from my time in New Jersey.   My friends have been a constant source of support and encouragement during my time in grad school, greatly enriching my life outside of academia, including of course S. Thais, V. Hallock, B. Dewenter, A. Bacoyanis, A. Johnson, S. Boutwell, L. Walton, N. Meyers, R. Parker, B. Whitney, N. Trask, L. Jia, E. Solomon, L. Roesler, M. Montgomery, E. Knox, L. Appel, L. Akers, and so many others!

I thank the Jim, Susan, Margaret, and Monica Hershberger, and the Cassidy family, D. Ruskaup, D. Steger, M. Christman, and all lifelong friends of our family.  And of course, I thank my own family; aunts, uncles, and cousins, and especially my grandpa James Q. Beatty.

My work benefited immensely from open source projects and developers, including the makers and maintainers of \LaTeX, Python, Numpy, Tensorflow, Git, the Library Genesis Project, and Sci-Hub, among others.  

I credit my cat Bobo with all typos: she would happily spend as much time on my laptop as I do.  And of course, many thanks go to Elizabeth Crites.
